---
title: "GGPLOT ENRS"
output:
  html_document:
    df_print: paged
---

Import the libraries.


```{r}
library(tidyverse)
```


We're going to start with our results2018 dataframe and then create a list of just candidates for the N.C. state House.

OK, so let's get just the state House races.

```{r}
NCHouse2018 <- results2018 %>%
        filter(
                str_detect(contest_name,"NC HOUSE OF REPRESENTATIVES*")
                ) %>%
         group_by(contest_name, choice, choice_party) %>%
        summarize(Votes=sum(total_votes))
```

```{r}
ggplot( NCHouse2018 ) + geom_col( mapping = aes(x=choice, y=Votes) )
```


What is "geom_col" ?

```{r}
?geom_col
```




Anyway, that chart is pretty useless. Let's just look at one contest.
```{r}
NCHouse2018 %>% 
        filter(contest_name == "NC HOUSE OF REPRESENTATIVES DISTRICT 003") %>%
        ggplot() + #we don't need to use the dataframe as the argument here beacuse we are piping it in from the first line in this R statement
        geom_col( mapping = aes(x=choice, y=Votes) )
```

But if you want to put it in order, you need to use the reorder() function in the value you assign to the x-axis.

```{r}
NCHouse2018 %>% 
        filter(contest_name == "NC HOUSE OF REPRESENTATIVES DISTRICT 003") %>%
        ggplot() + #we don't need to use the dataframe as the argument here beacuse we are piping it from the first line in this R statement
        geom_col( mapping = aes(x=reorder(choice, Votes), y=Votes) )
```

Least to most probably makes more sense to folks. So put a minus sign before the field you want to order by.

```{r}
NCHouse2018 %>% 
        filter(contest_name == "NC HOUSE OF REPRESENTATIVES DISTRICT 003") %>%
        ggplot() + #we don't need to use the dataframe as the argument here beacuse we are piping it in from the first line in the R statement
        geom_col( mapping = aes(x=reorder(choice, -Votes), y=Votes) )
```


Now let's change to a different question....

Did one party have a lot more candidates running? (For example, in unopposed races?)

To graph the number of candidates from each party we can switch from the geom_col() function to the geom_bar().

The geom_bar() function by default does a count of the number of rows in each category for whatever field you select. So let's look at the party of each candidate, which in this dataset is the column called `"choice Party".

```{r}
ggplot(data = NCHouse2018) +
        geom_bar(mapping = aes(x=choice_party))
```

OK, but what if you want to just focus on that relatively small difference between Democratic and Republican candidates? We can filter the data right here inside ggplot.

```{r}
ggplot(
        data = 
                filter(NCHouse2018,
                       choice_party %in% # we are using %in% because we are looking for more than one value in the same column.
                               c("DEM", "REP")
                       )
        ) +
        geom_bar(
                mapping = 
                        aes(
                                x=choice_party
                                )
                )
```

Hmmm.. maybe it would be easier to see when expressed as a percentage...

What about looking at what percentage of all candidates that are in each party? 

You can explicitly turn the y axis of a bar chart into proportions using "y=..prop.." and "group=1" 

(Why? Because I said so...?)

```{r}
ggplot(
        data = 
                filter(NCHouse2018,
                       choice_party %in%
                               c("DEM", "REP")
                       )
        ) +
        geom_bar(
                mapping = 
                        aes(
                                x=choice_party,
                                y= ..prop.., #this is the new line
                                group = 1 #so is this
                                )
                )
```

So... is there REALLY a difference between geom_col() and geom_bar()? They look to me like they draw pretty much the same picture. 

Well, let's try it.

Remember this?

```{r}
#NCHouse2018 %>% 
#        filter(contest_name == "NC HOUSE OF REPRESENTATIVES DISTRICT 003") %>%
#        ggplot() + 
#        geom_col( mapping = aes(x=choice, y=Votes) )
```

What if we changed it to geom_bar() ?(... and dropped the y argument...)

```{r}
NCHouse2018 %>% 
        filter(contest_name == "NC HOUSE OF REPRESENTATIVES DISTRICT 003") %>%
        ggplot() + 
        geom_bar( mapping = aes(x=choice) )
```

Well this is dumb. Just a histogram that tells us we have one value for each of our candidates. We need to "weight" it... which means (in this case anyway) tell R the column you want to use as values.

```{r}
NCHouse2018 %>%
        filter(contest_name == "NC HOUSE OF REPRESENTATIVES DISTRICT 003") %>%
        ggplot() + 
        geom_bar( mapping = aes(x=choice, weight=Votes))
```


So... "weight=" is kinda like "y=" ... not sure why the world needs both, but whatever... 

Basically, use geom_bar when you want to chart the count of the number of rows in different categories. And use geom_col when you want to use the actual values in a column as the y-axis.


#This is probably the end of Oct. 21

# Probably begin Oct. 23 here


Let's color by party... 
```{r}
NCHouse2018 %>% 
        filter(contest_name == "NC HOUSE OF REPRESENTATIVES DISTRICT 003") %>%
        ggplot() + 
        geom_bar( mapping = aes(x=choice, 
                                weight=Votes,
                                fill=choice_party)) #this is the new stuff
```

R has default colors it will use, but you can also force it to use certain colors. Here is just one of *many* ways you can do that.

```{r}
NCHouse2018 %>% 
filter(contest_name == "NC HOUSE OF REPRESENTATIVES DISTRICT 003") %>% 
ggplot() + 
        geom_bar( mapping = aes(x=choice, 
                                weight=Votes,
                                fill=choice_party)) +
        scale_fill_manual(values=c("red", "blue", "black")) #In this case, the function goes outside and after the geom_bar() function.
```

Not quite. I guess the colors have to be in the same order as the legend (which in this case is alpha order. But there are ways to force the order to change.) 

```{r}
NCHouse2018 %>% 
filter(contest_name == "NC HOUSE OF REPRESENTATIVES DISTRICT 003") %>% 
ggplot() + 
        geom_bar( mapping = aes(x=choice, 
                                weight=Votes,
                                fill=choice_party)) +
        scale_fill_manual(values=c("blue", "black", "red")) #this is the new line
```

Let's make it pretty order... 
```{r}
NCHouse2018 %>% 
filter(contest_name == "NC HOUSE OF REPRESENTATIVES DISTRICT 003") %>% 
ggplot() + 
        geom_bar( mapping = aes(x=reorder(choice, -Votes), #change this line
                                weight=Votes,
                                fill=choice_party)) +
        scale_fill_manual(values=c("blue", "black", "red")) 
```

How many key strokes would it take you to make a chart like this for all 120 state House races? ... 

```{r}
NCHouse2018 %>% 
filter(contest_name == "NC HOUSE OF REPRESENTATIVES DISTRICT 120") %>% 
ggplot() + 
        geom_bar( mapping = aes(x=reorder(choice, -Votes),
                                weight=Votes,
                                fill=choice_party)) +
        scale_fill_manual(values=c("blue", "red")) 
```



Let's look at Appeals Court seats to demonstrate stacked bars... 

```{r}
results2018 %>% 
        filter(str_detect(contest_name,"NC COURT OF APPEALS JUDGE SEAT")) %>%
        group_by(contest_name, choice, choice_party) %>%
        summarize(Votes=sum(total_votes)) %>%
        ggplot() + 
        geom_col( mapping = aes(x=contest_name, 
                                y=Votes,
                                fill=choice_party) #instead of explictly setting colors, we can use another column (this time "choice Party") to determine the colors.
                  )
```

Is it easier to read next to each other?

If you want them side-by-side instead of stacked then you just add position = "dodge" to the aes() function

```{r}
results2018 %>% 
        filter(str_detect(contest_name,"NC COURT OF APPEALS JUDGE SEAT*")) %>%
        group_by(contest_name, choice, choice_party) %>%
        summarize(Votes=sum(total_votes)) %>%
        ggplot() + 
        geom_col( mapping = aes(x=contest_name, 
                                y=Votes,
                                fill=choice_party
                                ),
                  position = "dodge"
                  )
```

We will come back and make this prettier later, but for now we are going to move on to scatterplots. (and,really, to illustrate some more aesthetic settings that aren't relevant to bar charts).


Remember the very first bar chart we made?

```{r}
#ggplot( NCHouse2018 ) + geom_col( mapping = aes(x=choice, y=Votes) )
```

What do you think is the one small change we would need to make to turn it in to a scatterplot?


```{r}
ggplot( NCHouse2018 ) + geom_point( mapping = aes(x=choice, y=Votes) )
```

That's a pretty dumb scatterplot.

I wonder if we can make it less dumb if we add color.

```{r}
ggplot( NCHouse2018 ) +
        geom_point( 
                mapping = 
                        aes(x=choice,
                            y=Votes,
                            color=choice_party #set the color of the points to the values in the choice Party column
                            )
                )
```

Look at that, 9 of the 10 candidates with the most votes were all Democrats. Sup with that?

Anyway, we're going to start making some ugly graphics in order to just demonstrate some levers you can play with. This time we will set the size of the points to the values in the ElectionDayPct column (which we will create -- temporarily -- with good ol' group_by and summarize).

```{r}
results2018 %>% 
        filter(str_detect(contest_name,"NC HOUSE OF REPRESENTATIVES*")) %>%
        group_by(contest_name, choice, choice_party) %>%
        summarize(TotalVotes=sum(total_votes), 
                  ElectionDayPCT = sum(election_day)/sum(total_votes),
                  NotElectionDayPCT = (sum(total_votes) - sum(election_day)) / sum(total_votes) ) %>%
ggplot() +
        geom_point( 
                mapping = 
                        aes(x=choice,
                            y=TotalVotes,
                            color=choice_party,
                            size = ElectionDayPCT # here is the line that controls the size of the dots
                            )
                )
```

... And we can change the transparency of the points by using the "alpha" argument and letting its value be controled by a column that has numerical values. In this case, the "TotalVotes" column.

```{r}
results2018 %>% 
        filter(str_detect(contest_name,"NC HOUSE OF REPRESENTATIVES*")) %>%
        group_by(contest_name, choice, choice_party) %>%
        summarize(TotalVotes=sum(total_votes), 
                  ElectionDayPCT = sum(election_day)/sum(total_votes),
                  NotElectionDayPCT = (sum(total_votes) - sum(election_day)) / sum(total_votes) ) %>%
ggplot() +
        geom_point( 
                mapping = 
                        aes(x=choice,
                            y=TotalVotes,
                            color=choice_party,
                            size = ElectionDayPCT,
                            alpha = TotalVotes
                            )
                )
```

We can use a column with categorical values to set the shape of the points... 

```{r}
results2018 %>% 
        filter(str_detect(contest_name,"NC HOUSE OF REPRESENTATIVES") ) %>%
        group_by(contest_name, choice, choice_party) %>%
        summarize(TotalVotes=sum(total_votes), 
                  ElectionDayPCT = sum(election_day)/sum(total_votes),
                  NotElectionDayPCT = (sum(total_votes) - sum(election_day)) / sum(total_votes) ) %>%
ggplot() +
        geom_point( 
                mapping = 
                        aes(x=choice,
                            y=TotalVotes,
                            color=choice_party,
                            size = ElectionDayPCT,
                            shape = contest_name #new line is here.
                            )
                )
```

Aha. That's a useful warning. Probably just generally a good idea to realize that visualizing more than a few categories at a time becomes pretty useless. 
We can get rid of the legend to make it more readable... 

```{r}
results2018 %>% 
        filter(str_detect(contest_name,"NC HOUSE OF REPRESENTATIVES*") ) %>%
        group_by(contest_name, choice, choice_party) %>%
        summarize(TotalVotes=sum(total_votes), 
                  ElectionDayPCT = sum(election_day)/sum(total_votes),
                  NotElectionDayPCT = (sum(total_votes) - sum(election_day)) / sum(total_votes) ) %>%
ggplot() +
        geom_point( 
                mapping = 
                        aes(x=choice,
                            y=TotalVotes,
                            color=choice_party,
                            size = ElectionDayPCT,
                            shape = contest_name
                            ),
                show.legend=FALSE) # use this argument, outside of aes() but inside geom_point in this case, to not display any legend.
                
```

Oh yeah, you can make all KINDS of ugly visualizations.


For now, let's switch things up and try to see if we can make a graph that shows all votes for Republican House candidates vs. all votes for Democratic House candidates. You try ... 

```{r}

```

#Here is where Oct. 23 is likely to end.


# We will probably begin Oct. 28 here. 


We can fix our y axis labels. But we have to install the "scales" library first. Scales is just a package that helps make numbers pretty. https://www.tidyverse.org/articles/2018/08/scales-1-0-0/

One of the things we can do is put commas into the numbers. (We can also do "$" and "%" but you can do that later by checking out the link above.)


```{r}

library(scales)
```


Here is our visualization without pretty axis labels... 

```{r}
results2018 %>% 
        filter(str_detect(contest_name,"NC HOUSE OF REPRESENTATIVES*")) %>%
        mutate(Party = choice_party) %>%
        group_by(Party) %>%
        summarize(TotalVotes=sum(total_votes)) %>%
        ggplot(
                aes(x=reorder(Party, -TotalVotes),
                    y=TotalVotes)
                ) +
                geom_col()


```

... And this just adds commas to the numbers on the y axis....

```{r}
results2018 %>% 
        filter(str_detect(contest_name,"NC HOUSE OF REPRESENTATIVES*")) %>%
        mutate(Party = choice_party) %>%
        group_by(Party) %>%
        summarize(TotalVotes=sum(total_votes)) %>%
        ggplot(
                aes(x=reorder(Party, -TotalVotes),
                    y=TotalVotes)
                ) +
                geom_col() +
                scale_y_continuous(labels=comma) #this is the new stuff.


```

Let's make the labels pretty.

Here we will use the labs() function and three arguments to set the text for the "title" of the chart as welll as the labels for the x and y axes.

```{r}
results2018 %>% 
        filter(str_detect(contest_name,"NC HOUSE OF REPRESENTATIVES*")) %>%
        mutate(Party = choice_party) %>%
        group_by(Party) %>%
        summarize(TotalVotes=sum(total_votes)) %>%
        ggplot(
                aes(x=reorder(Party, -TotalVotes),
                    y=TotalVotes)
                ) +
                geom_col() +
                scale_y_continuous(labels=comma) +
        labs( #this is the new stuff
                title="Total Votes Cast By Party, N.C. House Races 2018",
                x="Party",
                y="Votes for all candidates"
                )

```

ggplot() does allow you to fine tune a lot of elements of the chart, but not without some hassle. So folks have created "themes" that allow you to quickly make some common changes to the graphic.

Let's do some cool themes. Some -- like "theme_light()" are built in... 

```{r}
results2018 %>% 
        filter(str_detect(contest_name,"NC HOUSE OF REPRESENTATIVES*")) %>%
        mutate(Party = choice_party) %>%
        group_by(Party) %>%
        summarize(TotalVotes=sum(total_votes)) %>%
        ggplot(
                aes(x=reorder(Party, -TotalVotes),
                    y=TotalVotes)
                ) +
                geom_col() +
                scale_y_continuous(labels=comma) +
        labs( 
                title="Total Votes Cast By Party, N.C. House Races 2018",
                x="Party",
                y="Votes for all candidates"
                ) +
        theme_light() #this is the new stuff

```


Or, we can add themes that other fine people have made... like fivethirtyeight

http://www.ggplot2-exts.org/ggthemes.html

https://github.com/rgknight/theme_fivethirtyeight



```{r}

install.packages("ggthemes")

library("ggthemes")
```

```{r}
results2018 %>% 
        filter(str_detect(contest_name,"NC HOUSE OF REPRESENTATIVES*")) %>%
        mutate(Party = choice_party) %>%
        group_by(Party) %>%
        summarize(TotalVotes=sum(total_votes)) %>%
        ggplot(
                aes(x=reorder(Party, -TotalVotes),
                    y=TotalVotes,
                    )
                ) +
                geom_col() +
                scale_y_continuous(labels=comma) +
        labs( 
                title="Total Votes Cast By Party, N.C. House Races 2018",
                x="Party",
                y="Votes for all candidates"
                ) +

        theme_fivethirtyeight() #this is the new stuff

```


The ggthemes package also includes a function that allows you to fill your columns with the colors used by FiveThirtyEight (but only three colors...)

```{r}
results2018 %>% 
        filter(str_detect(contest_name,"NC HOUSE OF REPRESENTATIVES*")) %>%
        mutate(Party = choice_party) %>%
        group_by(Party) %>%
        summarize(TotalVotes=sum(total_votes)) %>%
        ggplot()+
                geom_bar(
                  mapping = 
                    aes(
                      x=reorder(Party, -TotalVotes),
                      weight=TotalVotes,
                      fill=Party)) +
                  scale_y_continuous(labels=comma) +
        labs( 
                title="Total Votes Cast By Party, N.C. House Races 2018",
                x="Party",
                y="Votes for all candidates"
                ) +

        theme_fivethirtyeight() +
        scale_fill_fivethirtyeight(7) #this is a separate function from theme_fivethirtyeight()





```

This does tell a less precise story, but you can combine the minor party candidates into one group called "Other"

```{r}
results2018 %>% 
        filter(str_detect(contest_name,"NC HOUSE OF REPRESENTATIVES*")) %>%
  #in this next mutate statement we both rename the column and group the "other" parties into one category.
        mutate(Party = case_when(choice_party %in% c("GRE","CST","LIB","UNA",NA) ~ "OTHER",
                  TRUE ~ as.character(choice_party)
                  )
        ) %>%
        group_by(Party) %>%
        summarize(TotalVotes=sum(total_votes)) %>%
        ggplot()+
                geom_bar(
                  mapping = 
                    aes(
                      x=reorder(Party, -TotalVotes),
                      weight=TotalVotes,
                      fill=Party)) +
                  scale_y_continuous(labels=comma) +
        labs( 
                title="Total Votes Cast By Party, N.C. House Races 2018",
                x="Party",
                y="Votes for all candidates"
                ) +

        theme_fivethirtyeight() +
        scale_fill_fivethirtyeight() 
```

We will learn how to match the colors to the correct column later.