---
title: "Intro to ggplot"
output:
  html_document:
    df_print: paged
---

Import the libraries

```{r}
library(tidyverse)
```
Read in the data
```{r}
#tabula_ncaa_football_combined <- read_csv("570-Fall91-Sports - R/data/tabula-ncaa_football_combined.csv")

```

Chart the number of people who attended the games.

```{r}
ggplot( data=tabula_ncaa_football_combined ) + geom_col( mapping = aes(x=School, y=Attend) )
```

What is "geom_col" ?

```{r}
?geom_col
```


Well, this is giving us twice as many schools as I would expect. Why?

Oh, let's just look at 2017 and not both years. We can add a filter() function to the right side of the data argument.

```{r}
ggplot( data=filter(tabula_ncaa_football_combined, Year==2017 ) ) + geom_col( mapping = aes(x=School, y=Attend) )
```

That's a little better, but maybe just the top 10?

Find the 10 schools with the most attendance in 2018. (Note: top_n must go at the end.)

First we have to make it ... 

```{r}
tabula_ncaa_football_combined %>% filter(Year==2018) %>% arrange(desc(Attend)) %>% select(School, Attend) %>% top_n(10) 
```

.... then, if we like it, save it to a data frame... 

```{r}
top10 <- tabula_ncaa_football_combined %>% filter(Year==2018) %>% arrange(desc(Attend)) %>% select(School, Attend) %>% top_n(10) 
```

Now graph this subset of all schools.

```{r}
ggplot( data=top10 ) + geom_col( mapping = aes(x=School, y=Attend) )
```

But if you want to put it in order, you need to use the reorder() function in the value you assign to the x-axis.

```{r}
ggplot( data=top10 ) + geom_col( mapping = aes(x=reorder(School,Attend), y=Attend) )
```

Least to most probably makes more sense to folks. So put a minus sign before the field you want to order by.

```{r}
ggplot( data=top10 ) + geom_col( mapping = aes(x=reorder(School,-Attend), y=Attend) )
```


That's enough now with looking att just the raw attendance. I wonder how much the number of games played affects the attendance.

To look at how many teams played different numbers of games we can switch from the geom_col() function to the geom_bar().

The geom_bar() function by default does a count of the number of rows in each category for whatever field you select. So let's look at the number of games (as a category, not an actual value), which in this dataset is the column called "G".

```{r}
ggplot( data=tabula_ncaa_football_combined ) + geom_bar( mapping = aes(x=G) )
```

How about just 2017?

```{r}
ggplot( data=filter(tabula_ncaa_football_combined, Year==2017 ) ) + geom_bar( mapping = aes(x=G) )
```

What about looking at what percentage of teams had each number of games? 

You can explicitly turn the y axis of a bar chart into proportions using "y=..prop.." 

(Why? Because I said so...)

```{r}
ggplot( data=filter(tabula_ncaa_football_combined, Year==2017 ) ) + geom_bar( mapping = aes(x=G, y=..prop..) )
```


So... is there REALLY a difference between geom_col() and geom_bar()? They look to me like they draw pretty much the same picture. 

Well, let's try it.

Remember this?
```{r}
# ggplot( data=top10 ) + geom_col( mapping = aes(x=School, y=Attend) )
```

What if we changed it to geom_bar() ? (... and dropped the y argument...)

```{r}
ggplot( data=top10 ) + geom_bar( mapping = aes(x=School) )
```

Well this is dumb. Just a histogram that tells us we have one value for each of our schools. We need to "weight" it... which means (in this case anyway) tell R the column you want to use as values. 

```{r}
ggplot( data=top10 ) + geom_bar( mapping = aes(x=School, weight = Attend) )
```

So... "weight=" is kinda like "y=" ... not sure why the world needs both, but whatever... 


#This is probably the end of Oct. 21

# Probably begin Oct. 23 here
You know, going back to that chart we made of the distribution of the number of teams that played different numbers of games, I wonder which teams played the most games.

Unfortunately, I don't remember what "the most number of games" was. Was it 7? 8? 22? No idea. 

Luckily, I let the computer remember things like this for me. I can tell R to give me only the rows where the value of the "G" column is equal to the largest value in the "G" column. 



```{r}
ggplot( 
        data=filter(
                tabula_ncaa_football_combined, 
                Year==2018, 
                G==max(G)) ) + #here is the good stuff
        geom_bar( 
                mapping = 
                        aes(
                                x=reorder(School, -Attend),
                                weight = Attend)
                )
```

Now why do you supposed UCF and Hawaii hosted more games than anyone else?

I wonder if those two schools ALWAYS host more games than anyone else.


```{r}
ggplot( 
        data=filter(
                tabula_ncaa_football_combined, 
                Year==2017, #just change the value of this argument
                G==max(G)) ) + 
        geom_bar( 
                mapping = 
                        aes(
                                x=reorder(School, -Attend),
                                weight = Attend)
                )
```

Hmmm... not the same at all. The Florida schools I understand, but Memphis?

Could we put both years on the same graph? And maybe differentiate them by using different colors?

Why, yes... yes we can.

```{r}
ggplot( 
        data=filter(
                tabula_ncaa_football_combined, # we dropped the year filter that had been here
                G==max(G)) ) + 
        geom_bar( 
                mapping = 
                        aes(
                                x=reorder(School, -Attend),
                                weight = Attend,
                                fill = Year) #One little new line.
                )
```


BUT WHY THE FUNKY LEGEND? 

Because these "years" are really cateogrical values, not numerical values. So either change the column datatype in the dataframe. Or wrap the values here in factor() ... (or "as.character()") function.

```{r}
ggplot( data=filter(tabula_ncaa_football_combined, G==max(G)) ) +
        geom_bar( mapping = aes(x=reorder(School, -Attend), weight = Attend, fill = factor(Year)) ) #here we wrap the Year column into the factor() function to (temporarily) change the datatype.
```
 
 Oh dear God! It looks like a gender neutral nursery.

We can force R to change colors. What I'm showing you here is a very short-term solution specific to this case. But it gives you the idea... 
```{r}
ggplot( data=filter(tabula_ncaa_football_combined, G==max(G)) ) +
        geom_bar( mapping = aes(x=reorder(School, -Attend), weight = Attend, fill = factor(Year)) ) + 
        scale_fill_manual(values=c("red", "blue")) #In this case, the function goes outside and after the geom_bar() function.
```


One last bar thing before going to scatterplots. Stacked bars.

```{r}

ggplot( data= tabula_ncaa_football_combined ) +
        geom_col( 
                mapping = aes(
                        x=reorder(School, -Attend),
                        y=Attend, 
                        fill = factor(Year) #instead of explictly setting colors, we can use another column (this time "Year") to determine the colors.
                ) 
        )
        
```

But I don't want 2018 on the bottom.

I have to "relevel" the "Year" factors.

The relevel() function takes two arguments -- the data you are using, and then the "upper" level.

```{r}

ggplot( data= tabula_ncaa_football_combined ) +
        geom_col( 
                mapping = aes(
                        x=reorder(School, -Attend),
                        y=Attend, 
                        fill = relevel(
                                factor(Year),
                                "2018")
                )
        )
```

If you want them side-by-side instead of stacked then you just add position = "dodge" to the aes() function

```{r}

ggplot( data= tabula_ncaa_football_combined ) +
        geom_col( 
                mapping = aes(
                        x=reorder(School, -Attend),
                        y=Attend, 
                        fill = relevel(
                                factor(Year), 
                                "2018")
                ),
                position = "dodge" #new stuff right here.
        )
```

We're going to leave the geom_col() and geom_bar() functions for a second to learn a new geom to build scatterplots (and,really, to illustrate some more aesthetic settings that aren't relevant to bar charts).

Remember the very first bar chart we made?

#ggplot( data=tabula_ncaa_football_combined ) + geom_col( mapping = aes(x=School, y=Attend) )

What do you think is the one small change we would need to make to turn it in to a scatterplot?

```{r}
ggplot( data=tabula_ncaa_football_combined ) + geom_point( mapping = aes(x=School, y=Attend) )
```

That's a pretty dumb scatterplot.

I wonder if we can make it less dumb if we add color.

```{r}
ggplot( data=tabula_ncaa_football_combined ) + 
        geom_point( mapping = aes(x=School, 
                                  y=Attend, 
                                  color=factor(Year), #this lets the "Year" column control the color of the dots
                                  )
        )
        
```

I fail to see a meaningful pattern. Let's continue our data exploration.

We already see the School name and attendance. But what if we made the dots bigger based on the number of games the teams played.. 

```{r}
ggplot( data=tabula_ncaa_football_combined ) + 
        geom_point( mapping = aes(x=School, 
                                  y=Attend, 
                                  color=factor(Year),
                                  size = G #this lets the "G" column control the size of the dots.
                                  )
        )
        
```

I can't think of any good reason to do this, but you can let a column control the opacity of the dots. 

```{r}
ggplot( data=tabula_ncaa_football_combined ) + 
        geom_point( mapping = aes(x=School, 
                                  y=Avg., 
                                  color=factor(Year),
                                  size = G,
                                  alpha = Attend #The value of the "Attend" column controls the opacity of the dots.
                                  )
        )
        
```

This is about to get seriously ugly, but in addition to color, size, and opacity (alpha) you can also pick a column that is used to control the shape of the dots.

```{r}
ggplot( data=tabula_ncaa_football_combined ) + 
        geom_point( mapping = aes(x=School, 
                                  y=Avg., 
                                  color=School,
                                  size = G,
                                  alpha = Attend,
                                  shape = factor(Year) #now the "Year" column is controlling the shape of the dots.
                                  )
        )
        
```

Well... When we set the color argument to "School" that gave us hundreds of colors that R wants to print in the legend. Let's get rid of that legend so we can see the shapes... 

```{r}
ggplot( data=tabula_ncaa_football_combined ) + 
        geom_point( mapping = aes(x=School, 
                                  y=Avg., 
                                  color=School,
                                  size = G,
                                  alpha = Attend,
                                  shape = factor(Year)
                                  ),
                            show.legend=FALSE) #this line gets rid of the legend.
        
```

#Here is where Oct. 23 is likely to end.


# We will probably begin Oct. 28 here. 
Now we can fix our y axis labels. But we have to install the "scales" library first. Scales is just a package that helps make numbers pretty. https://www.tidyverse.org/articles/2018/08/scales-1-0-0/


One of the things we can do is put commas into the numbers. (We can also do "$" and "%" but you can do that later by checking out the link above.)

```{r}

library(scales)

ggplot(top10, 
       aes(
               x=reorder(School, -Attend), 
               weight=Attend)) +
        geom_bar() +
        scale_y_continuous(labels=comma) #this is the new stuff.

```

Let's make the labels pretty.

Here we will use the labs() function and three arguments to set the text for the "title" of the chart as welll as the labels for the x and y axes.

```{r}
ggplot(top10, 
       aes(
               x=reorder(School, -Attend), 
               weight=Attend)
       ) +
        geom_bar() +
        scale_y_continuous(labels=comma) +
        labs( #this is the new stuff
                title="Top 10 Football Programs By Attendance, 2018",
                x="School",
                y="Attendance"
                )
```


ggplot() does allow you to fine tune a lot of elements of the chart, but not without some hassle. So folks have created "themes" that allow you to quickly make some common changes to the graphic.

Let's do some cool themes. Some -- like "theme_light()" are built in... 
```{r}
ggplot(top10, 
       aes(
               x=reorder(School, -Attend), 
               weight=Attend)
       ) +
        geom_bar() +
        scale_y_continuous(labels=comma) +
        labs( 
                title="Top 10 Football Programs By Attendance, 2018",
                x="School",
                y="Attendance"
                ) +
        theme_light() #this is the new stuff
```

Or, we can add themes that other fine people have made... like fivethirtyeight

http://www.ggplot2-exts.org/ggthemes.html

https://github.com/rgknight/theme_fivethirtyeight



```{r}

install.packages("ggthemes")

library("ggthemes")
```


```{r}
ggplot(top10, 
       aes(
               x=reorder(School, -Attend), 
               weight=Attend)
       ) +
        geom_bar() +
        scale_y_continuous(labels=comma) +
        labs( 
                title="Top 10 Football Programs By Attendance, 2018",
                x="School",
                y="Attendance"
                ) +
        theme_light()  + 
        theme_fivethirtyeight() #this is the new stuff

```



How about making horizontal bars instead of vertical? We "flip the cords"...

```{r}
ggplot(top10, 
       aes(
               x=reorder(School, -Attend), 
               weight=Attend)
       ) +
        geom_bar() +
        scale_y_continuous(labels=comma) +
        labs( 
                title="Top 10 Football Programs By Attendance, 2018",
                x="School",
                y="Attendance"
                ) +
        theme_light()  + 
        theme_fivethirtyeight() + 
        coord_flip() #this is the new stuff
```




